---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Reproducible Research: Peer Assessment 1

=====================================================================

## Loading and preprocessing the data

#### Reading data
```{r reading_data, echo = TRUE}
data <- read.csv(unzip("activity.zip"))
str(data)
```

#### Preprocessing data
```{r preprocessing_data, echo=TRUE}
library(plyr)
summary_by_date <- ddply(data,~date,
                         summarise,
                         total_steps=sum(steps, na.rm=TRUE),
                         mean_steps_per_interval=mean(steps, na.rm=TRUE))

data$interval <- as.factor(data$interval)
summary_by_interval <- ddply(data,~interval,
                         summarise,
                         total_steps=sum(steps, na.rm=TRUE),
                         mean_steps=mean(steps, na.rm=TRUE))
```

#### Summary of new data tables:
```{r summary, echo=TRUE}
str(summary_by_date)
str(summary_by_interval)
```

## What is mean total number of steps taken per day?

#### Computing Mean and Median
```{r Mean_Median, echo=TRUE}
mean_of_steps_per_day <- round(mean(summary_by_date$total_steps, na.rm=TRUE), 2)
median_of_steps_per_day <- median(summary_by_date$total_steps, na.rm=TRUE)
```

* Mean of total number of steps each day:  `r mean_of_steps_per_day` 
* Median of total number of steps each day: `r median_of_steps_per_day`

#### Histogram of Total Number of Steps each Day
```{r hist_totalsteps, echo=TRUE, fig.height=10, fig.width=18}
hist <- hist(summary_by_date$total_steps, 
     main = "Histogram of Total Steps per Day", breaks=20,
     xlab = "Total Steps each Day",
     ylab = "Frequency (days)",
     ylim = c(0,10),
     col = "BLUE",
     xaxt = "n")
axis(side=1, at=seq(0, 22000, 1000), lab=seq(0, 22000, 1000))
abline(v=mean_of_steps_per_day, col="BLACK", lty=2, lwd=2)
abline(v=median_of_steps_per_day, col="BLACK", lty=3, lwd=2)
legend("topright", legend=c("Mean", "Median"), col=c("BLACK", "BLACK"), lty=c(2,3), lwd=2)
```

## What is the average daily activity pattern?

#### Finding interval with maximum steps on an average
```{r max_steps_interval, echo = TRUE}
max_steps <- round(max(summary_by_interval$mean), digits=2)
max_index <- which.max(summary_by_interval$mean)
max_interval <- summary_by_interval$interval[max_index]
max_interval <- as.numeric(as.character(max_interval))
```

* The interval with the maximum number of steps is the `r max_interval`th interval with `r max_steps` steps everyday on an average

#### Time Series Plot : Average Number of Steps(y) Vs. Interval(x)

```{r timeseries, echo = TRUE, fig.height=10, fig.width=18}
plot(as.numeric(summary_by_interval$interval), summary_by_interval$mean, 
     type = "l", 
     col = "BLUE",
     main = "Average Steps per Interval", 
     xlab = "Interval",
     ylab = "Average Number of Steps (over all days)",
     xaxt = "n", 
     yaxt = "n")
axis(side=1, at=seq(1,288,4), labels=summary_by_interval$interval[seq(1,288,4)])
axis(side=2, at=seq(0,250,15), labels=seq(0,250,15))
abline(v=max_index, lwd=1, col="BLACK", lty=2)
abline(h=max_steps, lwd=1, col="BLACK", lty=2)
text(x=max_index+10, y=max_steps+4, labels=paste("(",max_interval,",",max_steps,")"))
```

## Imputing missing values

#### Computing total number of incomplete cases in the dataset
```{r incompletecases, echo = TRUE}
no_of_incomplete_cases <- nrow(data) - sum(complete.cases(data))
```
* Number of incomplete rows are `r no_of_incomplete_cases` out of `r nrow(data)` observations

#### Creating new data set with filled in values
* Strategy :
  * Replace NA values of incomplete cases with Mean of the number of steps for that interval 
  * If Mean of the number of steps is NaN, it will again result in incomplete cases. These are replaced by 0
```{r newdataset, echo = TRUE}
newdata <- data
newdata[!complete.cases(newdata),]$steps <- round(summary_by_interval[!complete.cases(newdata),]$mean_steps,2)
newdata[!complete.cases(newdata),]$steps <- 0
```

#### Preprocessing newdata
```{r new_preprocessing_data, echo=TRUE}
new_summary_by_date <- ddply(newdata,~date,
                         summarise,
                         total_steps=sum(steps, na.rm=TRUE),
                         mean_steps_per_interval=mean(steps, na.rm=TRUE))

newdata$interval <- as.factor(newdata$interval)
new_summary_by_interval <- ddply(newdata,~interval,
                         summarise,
                         total_steps=sum(steps, na.rm=TRUE),
                         mean_steps=mean(steps, na.rm=TRUE))
```

#### Computing Mean and Median
```{r new_Mean_Median, echo=TRUE}
new_mean_of_steps_per_day <- round(mean(new_summary_by_date$total_steps, na.rm=TRUE), 2)
new_median_of_steps_per_day <- median(new_summary_by_date$total_steps, na.rm=TRUE)
```

* Mean of total number of steps each day:  `r new_mean_of_steps_per_day` 
* Median of total number of steps each day: `r new_median_of_steps_per_day`  
    * Mean shifts by `r -mean_of_steps_per_day + new_mean_of_steps_per_day` steps to the right
    * Median shifts `r -median_of_steps_per_day + new_median_of_steps_per_day` steps to the right
    * The distribution remains almost the same

#### Histogram of Total Number of Steps each Day using Filled-In Data
```{r new_hist_totalsteps, echo=TRUE, fig.height=10, fig.width=18}
newhist <- hist(new_summary_by_date$total_steps, 
     main = "Histogram of Total Steps per Day Using Filled In Data", breaks=20,
     xlab = "Total Steps each Day",
     ylab = "Frequency (days)",
     ylim = c(0,10),
     col = "BLUE",
     xaxt = "n")
axis(side=1, at=seq(0, 22000, 1000), lab=seq(0, 22000, 1000))
abline(v=new_mean_of_steps_per_day, col="BLACK", lty=2, lwd=2)
abline(v=new_median_of_steps_per_day, col="BLACK", lty=3, lwd=2)
legend("topright", legend=c("Mean", "Median"), col=c("BLACK", "BLACK"), lty=c(2,3), lwd=2)
```


## Are there differences in activity patterns between weekdays and weekends?
```{r adding_weekdays, echo=TRUE, fig.height=10, fig.width=18}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
weekends <- c("Saturday", "Sunday")
newdata$day_of_week <- weekdays(as.Date(newdata$date))

# add a weekend/weekday column to newdata
newdata[newdata$day_of_week %in% weekends,]$day_of_week <- "weekend"
newdata[newdata$day_of_week %in% weekdays,]$day_of_week <- "weekday"

# creating factor variable
newdata$day_of_week <- as.factor(newdata$day_of_week)

# summarising with respect to two factor variables
weekdata <- ddply(newdata, .(interval, day_of_week), 
                                      summarise,
                                      total_steps=sum(steps, na.rm=TRUE),
                                      mean_steps=round(mean(steps, na.rm=TRUE),2))  

# lattice panel plot
library(lattice)
xyplot(mean_steps ~ interval | day_of_week, 
       data = weekdata, 
       layout = c(1,2), 
       type = "l", 
       ylab = "Number of Steps",
       xlab = "Interval", 
       scales=list(
          x=list(
                  at=seq(1,288,10), 
                  labels=weekdata$interval[seq(1,576,20)]
    )))
```

* Inferences from the above plot :
  * The activity pattern remains somewhat similar with peaks and drops within the same interval range
  * The number of steps on an average per interval is mostly lesser during the weekend.
  * The increase in the number of steps is earlier during weekdays as compared to weekends i.e approximately 535th interval on weekdays and 635th interval on weekdays - around an hour later